---
title: "Getting Started with llmbiobench"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with llmbiobench}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(llmbiobench)
```

## What this package is for

If you've ever tried to compare how ChatGPT, Claude, and Gemini handle the
same bioinformatics task, you've probably run into the same problem: there's
no good way to keep the input data, the reference answer, and the grading
rules all in one place, let alone compare five models against the same
task without redoing the setup each time. That's what llmbiobench is for.

Every benchmark test lives in its own folder with a predictable layout:

```
my-task/
  data/          the input files a model gets to see
  prompt/        the instructions sent to the model
  ref_answer/    the correct answer, hidden from the model
  ref_script/    the reference code that produced that answer
  answer/        one subfolder per model, holding what it submitted
  task.json      the question, in machine-readable form
  eval.json      how to score it
```

The functions in this package build that folder, write the two JSON files,
run models against it, and check the results, so you don't have to do any
of it by hand.

## Setting up a task

Say you want to test whether a model can reverse-complement a DNA sequence.
Start by giving the task an id and creating its folder structure:

```{r}
work_dir <- tempfile("llmbiobench-vignette-")
dir.create(work_dir)

bench_folders("revcomp-demo", output_dir = work_dir)
list.files(file.path(work_dir, "revcomp-demo"))
```

Now describe the question itself. `bench_task()` writes this out as
`task.json`, the file a model (or an agent reading the folder directly)
will actually see:

```{r}
bench_task(
  "revcomp-demo",
  output_dir = work_dir,
  question = "Reverse-complement the DNA sequence AATTGGGACCAACACTAGA and write it to a FASTA file.",
  output_file = "sequence.fasta",
  output_description = "A FASTA file with one record containing the reverse complement."
)
```

Before you can grade anything you need a reference answer sitting in
`ref_answer/`. Here we just write the correct sequence by hand, but in
practice this is usually generated by a real script, which is what
`ref_script/` is for.

```{r}
writeLines(
  c(">reverse_complement", "TCTAGTGTTGGTCCCAATT"),
  file.path(work_dir, "revcomp-demo", "ref_answer", "sequence.fasta")
)
writeLines(
  "# reference implementation lives here, e.g. Biostrings::reverseComplement()",
  file.path(work_dir, "revcomp-demo", "ref_script", "solution.R")
)
```

`bench_eval()` ties the question, the reference files, and a scoring
guideline together into `eval.json`. At an interactive console you can
leave `expected_output` out and it will ask you to name each output file
one at a time, but from a script (this vignette included) you need to
pass it in directly, since there's no terminal for it to prompt at:

```{r}
bench_eval(
  "revcomp-demo",
  output_dir = work_dir,
  question = "Reverse-complement the DNA sequence AATTGGGACCAACACTAGA and write it to a FASTA file.",
  expected_output = list(list(
    file = "sequence.fasta",
    guidelines = "Sequence must match the reference exactly; case and line-wrap may differ."
  ))
)
```

## Adding a rubric

A single pass/fail on the final file is fine for simple tasks, but often
you also care about *how* the model got there: did it hardcode a path, did
it peek at the answer key, did it use the right approach? `bench_rubric()`
builds on `bench_eval()` and adds a weighted rubric on top, mixing
criteria that can be checked mechanically (no LLM required) with ones that
genuinely need a judge to read the code.

```{r}
bench_rubric(
  "revcomp-demo",
  output_dir = work_dir,
  question = "Reverse-complement the DNA sequence AATTGGGACCAACACTAGA and write it to a FASTA file.",
  expected_output = list(list(
    file = "sequence.fasta",
    guidelines = "Sequence must match the reference exactly; case and line-wrap may differ."
  )),
  custom_criteria = list(list(
    id = "uses_biological_complement_rule",
    criterion = "The submitted code applies the A/T, C/G complement rule rather than a naive character substitution.",
    weight = 3, score_type = "binary",
    evidence = "Inspect the submitted script."
  ))
)
```

The mechanical criteria (`mechanical_rubric_criteria()`) cover the boring
but important stuff: is there an output file at all, does it parse, does
it match the reference, does the script avoid reading `ref_answer/` or
`eval.json` directly. You get those for free every time.

## Sending the prompt templates along

Every task folder gets a copy of the standard prompt templates, with the
task id already substituted in:

```{r}
add_prompts("revcomp-demo", output_dir = work_dir, version = "_v2")
list.files(file.path(work_dir, "revcomp-demo", "prompt"))
```

## Running a model against it

This is the part that actually talks to an LLM, so it needs a real API
key and isn't run when this vignette is built. `bench_llm()` reads the
prompt, hands the model a sandboxed set of `read_file`/`write_file`/
`list_files` tools scoped to the task folder (it can see `data/` and
`task.json`, but not `ref_answer/`, `ref_script/`, or either eval file),
and lets it work until it submits an answer under `answer/<model_name>/`.

```{r, eval = FALSE}
bench_llm(
  "revcomp-demo",
  provider = "anthropic",
  model = "claude-opus-4-5",
  output_dir = work_dir
)
```

Run this once per provider/model you want to compare. Each one gets its
own subfolder under `answer/`, so nothing gets overwritten.

## Grading the answer

Once you have one or more submissions sitting in `answer/`, there are two
ways to grade them, and you'll usually want both.

`bench_verify()` is the mechanical pass: no model involved, just direct
comparison against the reference. It knows how to compare numeric
matrices, mixed string/numeric tables, FASTA sequences, and plain numeric
text files, and it also pre-checks whichever rubric criteria from
`eval_v2.json` don't need a judge. Let's simulate a submission and see
what it says:

```{r}
dir.create(file.path(work_dir, "revcomp-demo", "answer", "toy-model"), recursive = TRUE)
writeLines(
  c(">reverse_complement", "TCTAGTGTTGGTCCCAATT"),
  file.path(work_dir, "revcomp-demo", "answer", "toy-model", "sequence.fasta")
)
writeLines(
  "# writes the reverse complement using Biostrings",
  file.path(work_dir, "revcomp-demo", "answer", "toy-model", "solution.R")
)

result <- bench_verify(file.path(work_dir, "revcomp-demo"))
```

Everything `bench_verify()` can't settle on its own gets flagged
`NEEDS_JUDGE` rather than silently skipped, which is where `bench_judge()`
comes in. It's the mirror image of `bench_llm()`: instead of hiding the
reference materials, it hands the judge model read access to them (plus
the one candidate's submission, never anyone else's), and lets it write
a verdict to `answer/<model>/judge_result.json`. Like `bench_llm()`, it
needs real credentials, so it's not run here either.

```{r, eval = FALSE}
bench_judge(
  "revcomp-demo",
  provider = "anthropic",
  model = "claude-opus-4-5",
  candidate_model = "toy-model",
  output_dir = work_dir
)
```

## Running a whole suite at once

If you're comparing several models across several tasks, calling
`bench_llm()`/`bench_judge()`/`bench_verify()` by hand for every
combination gets old fast. `bench_run_suite()` loops over every task and
every model, runs all three steps, and writes one long-format CSV you can
sort, filter, or plot. A failure on any single task/model pair is recorded
as an error row rather than stopping the whole run.

```{r, eval = FALSE}
models <- data.frame(
  provider = c("anthropic", "openai"),
  model = c("claude-opus-4-5", "gpt-5")
)

bench_run_suite(
  test_ids = c("revcomp-demo"),
  models = models,
  judge_provider = "anthropic",
  judge_model = "claude-opus-4-5",
  output_dir = work_dir
)
```

## Skipping the setup entirely: importing real tasks

Writing a task by hand is fine for a one-off, but PromptBio-Bench already
has over two hundred real, expert-curated bioinformatics tasks sitting on
Hugging Face. If you've fetched one locally with
`download_promptbio_task_files()` (in `code/read_promptbio_huggingface.R`),
`import_promptbio_task()` drops it straight into the standard folder
layout, ready for `bench_llm()`:

```{r, eval = FALSE}
import_promptbio_task("a-1-1", output_dir = work_dir)
```

Its `task.json`/`eval.json` already match the schema `bench_task()` and
`bench_eval()` produce, so this is just a matter of copying files into
place, not translating anything.

```{r, include = FALSE}
unlink(work_dir, recursive = TRUE)
```
